A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-08-02 analysis/paper/paper.Rmd
Final adjustments before publishing pre-print
Minor edits

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data seperate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in surveys, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017; @nummedal1923]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.
